Vagueness, Noise, and Signaling

### 2015-04-03 paper/derivation_RDD/derivation_RDD.tex
derivation RDD minor edits

We derive the proposed \rdd as the mean field of an imitative revision
protocol, following the example 4.3.1 of
\citet{Sandholm2010:Population-Game}. Our derivation relies on the
usual assumptions about conditional imitation at the agent level, but
must add special assumptions about how noise affects revision
probabilities. In keeping with the idea that agents in evolutionary
models are largely myopic, we assume here that noise is real but that
agents are unaware of it. When agents perceive the actual expected
utility of their choice and of another agent's choice (as is usual in
the motivation of imitation-based revision protocols), we assume
additionally that agents confuse their possibly non-veridical
subjective representations of states with the actual ones (details
below).

\paragraph{Notation.} As previously in this paper, let
$\Sstrat(\mymessg{i} \probbar \mystate{o})$ be the probability that a
speaker randomly picked from the population chooses $\mymessg{i}$ if
the actual/objective state is $\mystate{o}$. This factors in the
possibility that the sampled speaker perceives the objective state
$\mystate{o}$ as some subjective state $\mystate{s}$. In contrast, let
$\Sstrat_s(\mymessg{i} \probbar \mystate{s})$ be the probability that
a randomly sampled speaker chooses $\mymessg{i}$ if she experiences
some objective state $\mystate{o}$ as $\mystate{s}$. This is then the
hypothetical behavior of the speaker population in a (possibly
counterfactual) world without perceptual confusion. It is also what
the speakers actually do from their subjective point of view, and it
is this that changes in behavior (e.g., by imitation) will change in
the first place. Changes to $\Sstrat_s$ will, of course, also imply
changes in $\Sstrat$, as the latter is the result of what actually
happens when agents would like to do behave as in $\Sstrat_s$ in a
noisy world, as described by Equation~XYZ.

Similarly, $\Rstrat(\mystate{o} \probbar \messg)$ is the probability
that a randomly sampled receiver actually realizes objective state
$\mystate{o}$. $\Rstrat_s(\mystate{s} \probbar \messg)$ is the
probability of a subjectively intended interpretation $\mystate{s}$
that a randomly sampled receiver would pick in a noise-free
world. Again, behavior changes cause changes in $\Rstrat_s$, which in
turn influence $\Rstrat$.

@@

\paragraph{Imitation under perceptual noise.} There are many
conceivable ways in which perceptual noise might influence the above
revision protocol. Our choice here is guided by theoretical and
practical considerations alike. We would like maintain the assumption
that agents in evolutionary models are not necessarily fully aware of
their surroundings. In particular, we assume here that agents are not
themselves aware of perceptual noise. Switching probabilities are
therefore to be defined in terms of actual expected utility (and not
subjective expectations about actual expected utility that take noise
into consideration). Moreover, we assume that agents who have
(unconscious) access to expected utilities for actual states
$\mystate{o}$ will confuse whatever subjective representation
$\mystate{s}$ they may have had during their revision opportunity for
the actual state $\mystate{o}$.

Concretely, suppose agent $A$ has a revision opportunity for objective
state $\mystate{o}$, and chooses $\mymessg{i}$ with probability
$\Sstrat(\mymessg{i} \probbar \mystate{o})$. Similarly, $B$ chooses
$\mymessg{j}$ with probability $\Sstrat(\mymessg{j} \probbar
\mystate{o})$. It may be that $A$ perceived $\mystate{o}$ as
$\mystate{s_A}$ and that $B$ perceived $\mystate{s_B}$, but if $A$ is
unaware of noise he would see expect and see no difference. A observes
(as usual) the actual expected utilities of $\mymessg{i}$ and
$\mymessg{j}$ for the actual state $\mystate{o}$. Consequently, he
updates his subjective behavior for $\mystate{o}$ (even if he has
actually chosen $\mymessg{i}$ since he perceived
$\mystate{s_A}$). This myopic revision behavior gives rise to expected
changes in the average subjective/noise-free behavior $\Sstrat_s$ as
follows:
\begin{align*}
  \dot{\Sstrat}_s(\mymessg{i} \probbar \state) & = \sum_j
  \Sstrat(\mymessg{j} \probbar \state) \cdot p_{ji} -  
  \Sstrat(\mymessg{i} \probbar \state) \cdot \sum_j p_{ij} \\
  & = \sum_j  \Sstrat(\mymessg{j} \probbar \state) \cdot
    \Sstrat(\mymessg{i} \probbar \state) \cdot [\EU(\mymessg{i}) -
    \EU(\mymessg{j})]_+ \\ 
   & \ \ \ \  - \Sstrat(\mymessg{i} \probbar \state) \cdot \sum_j
   \Sstrat(\mymessg{j} \probbar \state) \cdot [\EU(\mymessg{j}) -
    \EU(\mymessg{i})]_+ \\
   & = \Sstrat(\mymessg{i} \probbar \state) \cdot \left ( \sum_j
     \Sstrat(\mymessg{j} \probbar \state) \cdot \EU(\mymessg{i}) - \sum_j
     \Sstrat(\mymessg{j} \probbar \state) \cdot \EU(\mymessg{j})
   \right ) \\
   & = \Sstrat(\mymessg{i} \probbar \state) \cdot \left (
     \EU(\mymessg{i}) - \text{average EU at $\state$} \right )
\end{align*}
This is the (continuous-time) replicator dynamics for behavioral
strategies.

Similar arguments let us derive the receiver part. Suppose that agent
$A$ has a revision opportunity for choice point $\messg$. He and a
random other agent $B$ actually choose $\mystate{o_A}$ and
$\mystate{o_B}$ with probability $\Rstrat(\mystate{o_{A,B}} \probbar
\messg)$, even though they may have intended to realize different
interpretations $\mystate{s_A}$ and $\mystate{s_B}$ coming from
$\Rstrat_s(\mystate{o_{A,B}} \probbar \messg)$, but then
noise-perturbed. Agent $A$ observes the actual expected utility for
the actually realized interpretations and, by assumptions, believes
(myopically) that he had actually wanted to realize $\mystate{o_A}$
and that $B$ had wanted to realize $\mystate{o_B}$. Under these
assumptions, the derivation of the replicator dynamic for subjectively
intended (noise-free) receiver behavior $\Rstrat_s$ proceeds in
parallel with the above sender case.

>>>>

We derive the proposed \rdd as a mean dynamic of an imitative revision
protocol. For expository reasons we follow closely the example 4.3.1
of \citet{Sandholm2010:Population-Game}, but other possibilities
exist. Our derivation relies on the usual assumptions about
conditional imitation at the agent level, but must add special
assumptions about how noise affects revision probabilities. In keeping
with the idea that agents are of possibly limited cognitive
capacities, we assume here that noise is real but that agents are
unaware of it. When agents perceive the actual expected utility of
their choice and of another agent's choice (as is usual in the
motivation of imitation-based revision protocols), we assume
additionally that agents mistake their possibly non-veridical
subjective representations of states for the actual ones (details
below).

\paragraph{Notation.} As previously in this paper, let
$\Sstrat(\mymessg{i} \probbar \mystate{o})$ be the probability that a
speaker randomly picked from the population chooses $\mymessg{i}$ if
the actual/objective state is $\mystate{o}$. This factors in the
possibility that the sampled speaker perceives the objective state
$\mystate{o}$ as some subjective state $\mystate{s}$. In contrast, let
$\Sstrat_s(\mymessg{i} \probbar \mystate{s})$ be the probability that
a randomly sampled speaker chooses $\mymessg{i}$ if she experiences
some objective state $\mystate{o}$ as $\mystate{s}$. This is then also
the average sender behavior in a (possibly counterfactual) world
without perceptual confusion. It is also where behavioral changes
first take effect, to then percolate to $\Sstrat$ as well.

Similarly, $\Rstrat(\mystate{o} \probbar \messg)$ is the probability
that a randomly sampled receiver chooses interpretation
$\mystate{o}$. $\Rstrat_s(\mystate{s} \probbar \messg)$ is the
probability of a subjectively intended interpretation
$\mystate{s}$. Again, behavioral changes first target $\Rstrat_s$ but
cause changes in $\Rstrat$ as well.

@@

\paragraph{Imitation under perceptual noise.} There are many
conceivable ways in which perceptual noise might influence the above
revision protocol. Our choice here is guided by theoretical and
practical considerations alike. We would like maintain the common
assumption that agents are ideal optimizers and not necessarily fully
aware of their surroundings. In particular, we assume here that agents
are not aware of perceptual noise. Switching probabilities are
therefore defined in terms of actual expected utility (and not
subjective expectations that take noise into consideration). Moreover,
we assume that agents who have (unconscious) access to expected
utilities for actual states $\mystate{o}$ will confuse whatever
subjective representation $\mystate{s}$ they may have had during their
revision opportunity for the actual state $\mystate{o}$.

Concretely, suppose agent $A$ has a revision opportunity for objective
state $\mystate{o}$, and chooses $\mymessg{i}$ with probability
$\Sstrat(\mymessg{i} \probbar \mystate{o})$. Similarly, $B$ chooses
$\mymessg{j}$ with probability $\Sstrat(\mymessg{j} \probbar
\mystate{o})$. It may be that $A$ perceived $\mystate{o}$ as
$\mystate{s_A}$ and that $B$ perceived $\mystate{s_B}$, but since $A$
is unaware of noise he expects no difference. $A$ observes (as usual)
the actual expected utilities of $\mymessg{i}$ and $\mymessg{j}$ for
$\mystate{o}$. Consequently, he updates his subjective behavior for
$\mystate{o}$ (not being aware that the perceived $\mystate{s_A}$
might have been different). This myopic revision behavior gives rise
to expected changes in the subjective/noise-free choice probabilities
$\Sstrat_s(\mymessg{i} \probbar \state)$ as follows:
\begin{align*}
  \dot{\Sstrat}_s(\mymessg{i} \probbar \state) & = \sum_j
  \Sstrat(\mymessg{j} \probbar \state) \cdot p_{ji} -  
  \Sstrat(\mymessg{i} \probbar \state) \cdot \sum_j p_{ij} \\
  & = \sum_j  \Sstrat(\mymessg{j} \probbar \state) \cdot
    \Sstrat(\mymessg{i} \probbar \state) \cdot [\EU(\mymessg{i}) -
    \EU(\mymessg{j})]_+ \\ 
   & \ \ \ \  - \Sstrat(\mymessg{i} \probbar \state) \cdot \sum_j
   \Sstrat(\mymessg{j} \probbar \state) \cdot [\EU(\mymessg{j}) -
    \EU(\mymessg{i})]_+ \\
   & = \Sstrat(\mymessg{i} \probbar \state) \cdot \left ( \sum_j
     \Sstrat(\mymessg{j} \probbar \state) \cdot \EU(\mymessg{i}) - \sum_j
     \Sstrat(\mymessg{j} \probbar \state) \cdot \EU(\mymessg{j})
   \right ) \\
   & = \Sstrat(\mymessg{i} \probbar \state) \cdot \left (
     \EU(\mymessg{i}) - \text{average EU at $\state$} \right )
\end{align*}
This is the (continuous-time) replicator dynamics for behavioral
strategies.

Similar considerations motivate the receiver part. Suppose that agent
$A$ has a revision opportunity for choice point $\messg$. He and a
random other agent $B$ actually choose $\mystate{o_A}$ and
$\mystate{o_B}$ with probability $\Rstrat(\mystate{o_{A,B}} \probbar
\messg)$, even though they may have intended interpretations
$\mystate{s_A}$ and $\mystate{s_B}$. Agent $A$ observes the actual
expected utility and, by assumption, believes (myopically) that he had
wanted to realize $\mystate{o_A}$ and that $B$ had wanted to realize
$\mystate{o_B}$. Under these assumptions, the derivation of the
replicator dynamic for subjectively intended (noise-free) receiver
behavior $\Rstrat_s$ proceeds in parallel with the above sender case.
